Final analysis of Greenland narwhals trajectories with smoothSDE

### 2025-01-29 R/application/RCVM/fit_rcvm.Rmd
new results with shrunk land polygons

library(smoothSDE)
library(ggplot2)
library(dplyr)
library(plotly)
library(htmlwidgets)
library(mgcv)
library(here)
library(xtable)
library(sf)
library(gganimate) # animated trajectories
```

@@

@@

@@

@@

#Fixed measurement error
sigma_obs=0.03
H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_pre12))

#Fit baseline with measurement error estimated from the data
baseline_ctcrw0<- SDE$new(formulas = formulas,data = dataBE12,type = "CTCRW",
                          response = c("x","y"),
                          par0 = par0,
                          other_data=list("log_sigma_obs0"=log(0.05)),
                          fixpar=c("mu1","mu2"))
baseline_ctcrw0$fit()

@@

```{r results baseline_ctcrw0}

estimates_bas_ctcrw0=as.list(baseline_ctcrw0$tmb_rep(),what="Est")
std_bas_ctcrw0=as.list(baseline_ctcrw0$tmb_rep(),what="Std")

@@

```{r render estimates table baseline_ctcrw0}

# Function to extract estimate values into a data frame format
combine_to_df <- function(estimates, std_dev, param_name) {
  data.frame(
    Parameter = param_name,
    Estimate = unlist(estimates, use.names = FALSE),
    Std_Error = unlist(std_dev, use.names = FALSE)
  )
}

# Create data frames for each set
log_sigma_obs <- combine_to_df(estimates_bas_ctcrw0$log_sigma_obs, 
                               std_bas_ctcrw0$log_sigma_obs, "log_sigma_obs")
coeff_fe <- combine_to_df(estimates_bas_ctcrw0$coeff_fe,
                          std_bas_ctcrw0$coeff_fe, 
                          rownames(estimates_bas_ctcrw0$coeff_fe))

coeff_re <- combine_to_df(estimates_bas_ctcrw0$coeff_re,
                          std_bas_ctcrw0$coeff_re, 
                          rownames(estimates_bas_ctcrw0$coeff_re))

log_lambda <- combine_to_df(estimates_bas_ctcrw0$log_lambda,
                          std_bas_ctcrw0$log_lambda, 
                          rownames(estimates_bas_ctcrw0$log_lambda))

@@

```{r, results='asis',message=FALSE}
latex_table <- xtable(table_data, 
                      caption = "Parameter Estimates and Standard Errors")

# Print LaTeX table
print(latex_table, type = "latex", include.rownames = FALSE)
```
We can get confidence intervals for the intercept of each parameter and the random effects standard deviations.

```{r CI baseline_ctcrw0 parameters}

post_coeff_bas_ctrcw0=baseline_ctcrw0$post_coeff(n_post=10000)
post_par_bas_ctcrw0=list(
  "tau"=exp(post_coeff_bas_ctrcw0$coeff_fe[,"tau.(Intercept)"]),
  "nu"=exp(post_coeff_bas_ctrcw0$coeff_fe[,"nu.(Intercept)"]),
  "sigma_tau"=1/sqrt(exp(post_coeff_bas_ctrcw0$log_lambda[,"tau.s(ID)"])),
  "sigma_nu"=1/sqrt(exp(post_coeff_bas_ctrcw0$log_lambda[,"nu.s(ID)"])))

mean_par_bas_ctcrw0=lapply(post_par_bas_ctcrw0,mean)
sd_par_bas_ctcrw0=lapply(post_par_bas_ctcrw0,sd)
quant_par_bas_ctcrw0 <- lapply(post_par_bas_ctcrw0, 
                               quantile, probs = c(0.025, 0.975))

@@

```{r CI table baseline_ctcrw0 parameters, message=FALSE}

parameter_names <- names(mean_par_bas_ctcrw0)
estimates <- mean_par_bas_ctcrw0
conf_intervals <- sapply(parameter_names, function(param) {
  sprintf("$[%.2f; %.2f]$", quant_par_bas_ctcrw0[[param]][1], 
          quant_par_bas_ctcrw0[[param]][2])
})

table_data <- data.frame(
  Parameter = 
    c("$\\tau$", "$\\nu$","$\\sigma_{\\tau}$","$\\sigma_{\\nu}$"),
  Estimate = sprintf("$%.2f$", mean_par_bas_ctcrw0),
  CI = conf_intervals,
  stringsAsFactors = FALSE
)

```{r,results='asis',echo=FALSE}
xtab <- xtable(
  table_data,
  caption = "CTCRW baseline estimations with random effects",
  label = "table:ctcrw_baseline_estimations_with_random_effects"
)

print(
  xtab,type="latex",
  include.rownames = FALSE,
  sanitize.text.function = identity,  # Prevent escaping of LaTeX math symbols
  hline.after = c(-1, 0, nrow(table_data)),  
  table.placement="H"
)
```

@@

We now decide to introduce a rotational parameter $\omega$ in the SDE to allow the velocity to rotate depending on the narwhals position with respect to the shoreline.
We start by introducing the covariate \texttt{AngleNormal} in the new smooth parameter $\omega$ to understand how the animals may rotate depending on the direction of their movement with respect to the shoreline.

@@

#model formula
formulas <- list(mu1=~1,mu2=~1,
                 tau =~s(theta_translated,k=10,bs="cc")+s(ID,bs="re"),
                 nu=~s(ID,bs="re"),
                 omega=~s(theta_translated,k=10,bs="cs"))

#Fixed measurement error
sigma_obs=0.03
H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_pre12))

# Fit baseline with measurement error estimated from the data
baseline_racvm0<- SDE$new(formulas = formulas,data = dataBE12,type = "RACVM_SSM",response = c("x","y"),
                    par0 = par0,other_data=list("log_sigma_obs0"=log(0.05)),fixpar=c("mu1","mu2"))

@@

```{r results baseline_racvm0}

estimates_bas_racvm0=as.list(baseline_racvm0$tmb_rep(),what="Est")
std_bas_racvm0=as.list(baseline_racvm0$tmb_rep(),what="Std")
```

```{r render estimates table baseline_racvm0 }

# Function to extract estimate values into a data frame format
combine_to_df <- function(estimates, std_dev, param_name) {
  data.frame(
    Parameter = param_name,
    Estimate = unlist(estimates, use.names = FALSE),
    Std_Error = unlist(std_dev, use.names = FALSE)
  )
}

# Create data frames for each set
log_sigma_obs <- combine_to_df(estimates_bas_racvm0$log_sigma_obs, 
                               std_bas_racvm0$log_sigma_obs, "log_sigma_obs")
coeff_fe <- combine_to_df(estimates_bas_racvm0$coeff_fe, 
                          std_bas_racvm0$coeff_fe, 
                          rownames(estimates_bas_racvm0$coeff_fe))
coeff_re <- combine_to_df(estimates_bas_racvm0$coeff_re, 
                          std_bas_racvm0$coeff_re, 
                          rownames(estimates_bas_racvm0$coeff_re))
log_lambda <- combine_to_df(estimates_bas_racvm0$log_lambda, 
                          std_bas_racvm0$log_lambda, 
                          rownames(estimates_bas_racvm0$log_lambda))

@@

```{r, results='asis',message=FALSE}
latex_table <- xtable(table_data, 
                      caption = "Parameter Estimates and Standard Errors")

# Print LaTeX table
print(latex_table, type = "latex", include.rownames = FALSE)
```

```{r CI baseline_racvm0 parameters}

post_coeff_bas_racvm0=baseline_racvm0$post_coeff(n_post=10000)
post_par_bas_racvm0=list(
  "tau"=exp(post_coeff_bas_racvm0$coeff_fe[,"tau.(Intercept)"]),
  "nu"=exp(post_coeff_bas_racvm0$coeff_fe[,"nu.(Intercept)"]),
  "omega"=post_coeff_bas_racvm0$coeff_fe[,"omega.(Intercept)"],
  "sigma_tau"=1/sqrt(exp(post_coeff_bas_racvm0$log_lambda[,"tau.s(ID)"])),
  "sigma_nu"=1/sqrt(exp(post_coeff_bas_racvm0$log_lambda[,"nu.s(ID)"])))

mean_par_bas_racvm0=lapply(post_par_bas_racvm0,mean)
sd_par_bas_racvm0=lapply(post_par_bas_racvm0,sd)
quant_par_bas_racvm0 <- lapply(post_par_bas_racvm0, 
                               quantile, probs = c(0.025, 0.975))

```{r CI table baseline_racvm0 parameters, message=FALSE}

parameter_names <- names(mean_par_bas_racvm0)
estimates <- mean_par_bas_racvm0
conf_intervals <- sapply(parameter_names, function(param) {
  sprintf("$[%.2f; %.2f]$", quant_par_bas_racvm0[[param]][1], 
          quant_par_bas_racvm0[[param]][2])
})

table_data <- data.frame(
  Parameter = 
    c("$\\tau$", "$\\nu$","$\\omega$","$\\sigma_{\\tau}$","$\\sigma_{\\nu}$"),
  Estimate = sprintf("$%.2f$", mean_par_bas_racvm0),
  CI = conf_intervals,
  stringsAsFactors = FALSE
)

```{r,results='asis',echo=FALSE}
xtab <- xtable(
  table_data,
  caption = "racvm0 baseline estimations",
  label = "table:racvm0_baseline_estimations_with_random_effects"
)

print(
  xtab,type="latex",
  include.rownames = FALSE,
  sanitize.text.function = identity,  # Prevent escaping of LaTeX math symbols
  hline.after = c(-1, 0, nrow(table_data)),  
  table.placement="H"
)
```

@@

We add the effect of the distance to the shore through the covariate \texttt{Ishore}.
This model has a lot of parameters and takes around 15 minutes to fit. 
Indeed, we will need to fix the penalties in the tensor splines in advance to ensure convergence of the optimizer. The standard deviation of the random effects and the smoothing penalties for the splines in $\tau$ will still be estimated from the data and not fixed. To illustrate this we first fit a model where the smoothing penalties are all estimated from the data.

@@

#Fixed measurement error
sigma_obs=0.03
H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_pre12))

#parameters formulas
formulas <- list(mu1 = ~1 ,mu2 =~1,tau =~s(theta_translated,k=10,bs="cc")+s(ID,bs="re"),nu=~s(ID,bs="re"),
                   omega=~te(theta_translated,DistanceShore_translated,k=5,bs="cs"))

@@

```{r get results baseline_racvm1_unstable,message=FALSE}

estimates_bas_racvm1_unstable=as.list(baseline_racvm1_unstable$tmb_rep(),what="Est")
std_bas_racvm1_unstable=as.list(baseline_racvm1_unstable$tmb_rep(),what="Std")

estimates_bas_racvm1_unstable

```
```{r std random effects baseline_racvm1_unstable}

@@

We see that the estimate for the standard deviation $\sigma_\tau$ of the random effect for $\tau$ is very high while the estimate for the standard deviation $\sigma_\nu$ of the random effect for $\nu$ is very close to $0$. The smoothing penalties estimates are extremely high (the smoothing penalties are the exponential of \texttt{estimates\_bas\_racvm1\_unstable$log\_lambda}). This suggests that the optimization did not go well and numerical errors occurred. For this reason, we decide to fix the smoothing penalties to "reasonable" values in the tensor splines.

First, we get good values for the smoothing penalty by approximating the true surface with splines.

@@

fomega=function(cov_data,a=2,b=5,D0=0.5,D1=0.8,sigma_D=0.3,sigma_theta=pi/6){
  Dshore=cov_data$DistanceShore
  theta=cov_data$theta
  if (is.null(Dshore)){
    Dshore=1/cov_data$Ishore
  }

@@

# Splines smooth for omega -------------------- 
n <- 100000
SP_DF=5

#sample theta and DistanceShore points in the domain
theta <- runif(n,-pi,pi);
DistanceShore <- runif(n,0.02,7)
samples=data.frame(theta=theta,DistanceShore=DistanceShore)

#define grid of values
theta_v <- seq(-pi,pi,length=30)
Dshore_v<- seq(0.02,7,length=30)
pr <- data.frame(theta=rep(theta_v,30),DistanceShore=rep(Dshore_v,rep(30,30)))

@@

#fit with bivariate splines te
m1 <- gam(y~te(theta,DistanceShore,k=SP_DF))

@@

#initial parameters
par0 <- c(0,0,1,4,0)
init_lambda=c(exp(3),exp(3),1,exp(-12),exp(-5))

@@

#Fixed measurement error
sigma_obs=0.03
H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_pre12))

#parameters formulas
formulas <- list(mu1 = ~1 ,mu2 =~1,tau =~s(theta_translated,k=10,bs="cc")+s(ID,bs="re"),nu=~s(ID,bs="re"),
                   omega=~te(theta_translated,DistanceShore_translated,k=c(5,5),bs="cs"))

# Fit baseline
baseline_racvm1<- SDE$new(formulas = formulas,data = dataBE12,type = "RACVM_SSM",
                    response = c("x","y"),par0 = par0,other_data=list("log_sigma_obs0"=log(0.05)),
                    fixpar=c("mu1","mu2"))

@@

```{r results baseline_racvm1}

estimates_bas_racvm1=as.list(baseline_racvm1$tmb_rep(),what="Est")
std_bas_racvm1=as.list(baseline_racvm1$tmb_rep(),what="Std")
```

```{r render estimates table baseline_racvm1 }

# Create data frames for each set
log_sigma_obs <- combine_to_df(estimates_bas_racvm1$log_sigma_obs, 
                               std_bas_racvm1$log_sigma_obs, "log_sigma_obs")
coeff_fe <- combine_to_df(estimates_bas_racvm1$coeff_fe, 
                          std_bas_racvm1$coeff_fe, 
                          rownames(estimates_bas_racvm1$coeff_fe))
coeff_re <- combine_to_df(estimates_bas_racvm1$coeff_re, 
                          std_bas_racvm1$coeff_re, 
                          rownames(estimates_bas_racvm1$coeff_re))
log_lambda <- combine_to_df(estimates_bas_racvm1$log_lambda, 
                          std_bas_racvm1$log_lambda, 
                          rownames(estimates_bas_racvm1$log_lambda))

@@

```{r, results='asis',message=FALSE}
latex_table <- xtable(table_data, 
                      caption = "Parameter Estimates and Standard Errors")

# Print LaTeX table
print(latex_table, type = "latex", include.rownames = FALSE)
```

```{r CI baseline_racvm1 parameters}

post_coeff_bas_racvm1=baseline_racvm1$post_coeff(n_post=10000)
post_par_bas_racvm1=list(
  "tau"=exp(post_coeff_bas_racvm1$coeff_fe[,"tau.(Intercept)"]),
  "nu"=exp(post_coeff_bas_racvm1$coeff_fe[,"nu.(Intercept)"]),
  "omega"=post_coeff_bas_racvm1$coeff_fe[,"omega.(Intercept)"],
  "sigma_tau"=1/sqrt(exp(post_coeff_bas_racvm1$log_lambda[,"tau.s(ID)"])),
  "sigma_nu"=1/sqrt(exp(post_coeff_bas_racvm1$log_lambda[,"nu.s(ID)"])))

mean_par_bas_racvm1=lapply(post_par_bas_racvm1,mean)
sd_par_bas_racvm1=lapply(post_par_bas_racvm1,sd)
quant_par_bas_racvm1 <- lapply(post_par_bas_racvm1, 
                               quantile, probs = c(0.025, 0.975))

```{r CI table baseline_racvm1 parameters, message=FALSE}

parameter_names <- names(mean_par_bas_racvm1)
estimates <- mean_par_bas_racvm1
conf_intervals <- sapply(parameter_names, function(param) {
  sprintf("$[%.2f; %.2f]$", quant_par_bas_racvm1[[param]][1], 
          quant_par_bas_racvm1[[param]][2])
})

table_data <- data.frame(
  Parameter = 
    c("$\\tau$", "$\\nu$","$\\omega$","$\\sigma_{\\tau}$","$\\sigma_{\\nu}$"),
  Estimate = sprintf("$%.2f$", mean_par_bas_racvm1),
  CI = conf_intervals,
  stringsAsFactors = FALSE
)

```{r,results='asis',echo=FALSE}
xtab <- xtable(
  table_data,
  caption = "racvm1 baseline estimations",
  label = "table:racvm1_baseline_estimations_with_random_effects"
)

print(
  xtab,type="latex",
  include.rownames = FALSE,
  sanitize.text.function = identity,  # Prevent escaping of LaTeX math symbols
  hline.after = c(-1, 0, nrow(table_data)),  
  table.placement="H"
)
```
```{r display plots baseline_racvm1}

Ishore_values=c(1/0.25,1/2,1/10)
plots=lapply(1:length(Ishore_values),function(i){
  baseline_racvm1$plot_par(var="AngleNormal",covs=data.frame("ID"="Asgeir","Ishore"=Ishore_values[i]),
                         par_names=c("omega"),n_post=100)})
plots

@@

xmin=list("DistanceShore_translated"=0.2)
xmax=list("DistanceShore_translated"=2)

plots_bas_racvm1=baseline_racvm1$get_all_plots(baseline=NULL,
      xmin=xmin, xmax=xmax,show_CI="pointwise")

@@

Now, we would like to demonstrate how the Distance to shore can influence the smooth parameter $\omega$.
We plot the estimated smooth parameter $\omega$ as a function of the angle for different values of the Distance to shore.

```{r smooth omega baseline_racvm1}
plot1=plots_bas_racvm1$fe_omega_AngleNormal_q1_Ishore
plot2=plots_bas_racvm1$fe_omega_AngleNormal_q2_Ishore
plot4=plots_bas_racvm1$fe_omega_AngleNormal_q4_Ishore

@@

combined_plot <- ggplot() +
  geom_line(data=combined_omega, aes(x=x, y=y, color=Category, linetype=Category)) +
  geom_ribbon(data=combined_ci_omega, aes(x=x, ymin=ymin, ymax=ymax, fill=Category), alpha=0.1) +
  labs(color="Category", linetype="Category", x=expression(Theta), y=expression(omega)) +
  geom_vline(xintercept=c(-pi/2, pi/2), linetype="dashed", color="grey") +
  annotate("text", x = -pi/2-0.3, y = -5, label = expression(-pi/2), color = "red", size = 5) + 
  annotate("text", x = pi/2+0.2, y = -5, label = expression(pi/2), color = "red", size = 5) + 
  theme_minimal() +
  theme(
    axis.title.x = element_text(size=14),
    axis.title.y = element_text(size=14,angle=0,vjust=0.5, margin = margin(r = 10)),
    axis.text.y = element_text(size=12),
    axis.ticks.x = element_blank(),
    axis.text.x = element_blank(),
    legend.title = element_text(size=14),
    legend.text = element_text(size=12))

ggsave("omega_DistanceShore_levels.pdf", plot=combined_plot, path="baseline_racvm1", width=10, height=6, units="in")

combined_plot

@@

This model in indeed very complex and involves a large number of parameters given the amount of data points we have for the trajectories before exposure.
We can try a more parsimonious model to make estimation easier.

## Baseline with AngleNormal and Ishore in omega

We decide to remove the effect of \texttt{AngleNormal} on $\tau$ and only put covariates on $\omega$.

#initial parameters
par0 <- c(0,0,2,4,0)
init_lambda=c(exp(3),exp(3),exp(-12),exp(-5))

#parameters formulas
formulas <- list(mu1 = ~1 ,mu2 =~1,tau =~s(ID,bs="re"),nu=~s(ID,bs="re"),
                   omega=~te(theta_translated,DistanceShore_translated,k=5,bs="cs"))

```{r get results baseline_racvm2}

estimates_bas_racvm2=as.list(baseline_racvm2$tmb_rep(),what="Est")
std_bas_racvm2=as.list(baseline_racvm2$tmb_rep(),what="Std")

```{r render estimates table baseline_racvm2}
# Create data frames for each set
log_sigma_obs <- combine_to_df(estimates_bas_racvm2$log_sigma_obs, 
                               std_bas_racvm2$log_sigma_obs, "log_sigma_obs")
coeff_fe <- combine_to_df(estimates_bas_racvm2$coeff_fe, 
                          std_bas_racvm2$coeff_fe, 
                          rownames(estimates_bas_racvm2$coeff_fe))
coeff_re <- combine_to_df(estimates_bas_racvm2$coeff_re, 
                          std_bas_racvm2$coeff_re, 
                          rownames(estimates_bas_racvm2$coeff_re))
log_lambda <- combine_to_df(estimates_bas_racvm2$log_lambda, 
                          std_bas_racvm2$log_lambda, 
                          rownames(estimates_bas_racvm2$log_lambda))

```{r, results='asis',message=FALSE}
latex_table <- xtable(table_data, 
                      caption = "Parameter Estimates and Standard Errors")

# Print LaTeX table
print(latex_table, type = "latex", include.rownames = FALSE)
```
```{r CI baseline_racvm2 parameters}

post_coeff_bas_racvm2=baseline_racvm2$post_coeff(n_post=10000)
post_par_bas_racvm2=list(
  "tau"=exp(post_coeff_bas_racvm2$coeff_fe[,"tau.(Intercept)"]),
  "nu"=exp(post_coeff_bas_racvm2$coeff_fe[,"nu.(Intercept)"]),
  "omega"=post_coeff_bas_racvm2$coeff_fe[,"omega.(Intercept)"],
  "sigma_tau"=1/sqrt(exp(post_coeff_bas_racvm2$log_lambda[,"tau.s(ID)"])),
  "sigma_nu"=1/sqrt(exp(post_coeff_bas_racvm2$log_lambda[,"nu.s(ID)"])))

mean_par_bas_racvm2=lapply(post_par_bas_racvm2,mean)
sd_par_bas_racvm2=lapply(post_par_bas_racvm2,sd)
quant_par_bas_racvm2 <- lapply(post_par_bas_racvm2, 
                               quantile, probs = c(0.025, 0.975))

```{r CI table baseline_racvm2 parameters, message=FALSE}

parameter_names <- names(mean_par_bas_racvm2)
estimates <- mean_par_bas_racvm2
conf_intervals <- sapply(parameter_names, function(param) {
  sprintf("$[%.2f; %.2f]$", quant_par_bas_racvm2[[param]][1], 
          quant_par_bas_racvm2[[param]][2])
})

table_data <- data.frame(
  Parameter = 
    c("$\\tau$", "$\\nu$","$\\omega$","$\\sigma_{\\tau}$","$\\sigma_{\\nu}$"),
  Estimate = sprintf("$%.2f$", mean_par_bas_racvm2),
  CI = conf_intervals,
  stringsAsFactors = FALSE
)

```{r,results='asis',echo=FALSE}
xtab <- xtable(
  table_data,
  caption = "racvm2 baseline estimations",
  label = "table:racvm2_baseline_estimations_with_random_effects"
)

print(
  xtab,type="latex",
  include.rownames = FALSE,
  sanitize.text.function = identity,  # Prevent escaping of LaTeX math symbols
  hline.after = c(-1, 0, nrow(table_data)),  
  table.placement="H"
)
```
```{r display plots baseline_racvm2}

Ishore_values=c(1/0.25,1/2,1/5)
plots=lapply(1:length(Ishore_values),function(i){
  baseline_racvm2$plot_par(var="AngleNormal",covs=data.frame("ID"="Asgeir","Ishore"=Ishore_values[i]),
                         par_names=c("omega"),n_post=100)})
plots

>>>>

library(smoothSDE)
library(ggplot2)
library(dplyr)
library(plotly)
library(htmlwidgets)
library(mgcv)
library(here)
library(xtable)
library(sf)
library(gganimate) # animated trajectories
library(doParallel)
library(foreach)
source(file.path(here("R","simulation_study_CRCVM","CVM_functions.R"))) 
```

@@

dataBE24[dataBE24$DistanceShore_translated> D_up,"Ishore_translated"]=0
dataBE24[dataBE24$DistanceShore_translated< D_low,"Ishore_translated"]=1/D_low

@@

dataBE12[dataBE12$DistanceShore_translated> D_up,"Ishore_translated"]=0
dataBE12[dataBE12$DistanceShore_translated< D_low,"Ishore_translated"]=1/D_low

@@

dataAE[dataAE$DistanceShore_translated> D_up,"Ishore_translated"]=0
dataAE[dataAE$DistanceShore_translated< D_low,"Ishore_translated"]=1/D_low

@@

#Fit baseline with measurement error estimated from the data
baseline_ctcrw0<- SDE$new(formulas = formulas,data = dataBE12,type = "CTCRW",
                          response = c("x","y"),
                          par0 = par0,
                          other_data=list("H"=H),
                          fixpar=c("mu1","mu2"))
baseline_ctcrw0$fit()

@@

```{r define function to get raw estimates}
# Function to extract estimate values into a data frame format
combine_to_df <- function(estimates, std_dev, param_name) {
  data.frame(
    Parameter = param_name,
    Estimate = unlist(estimates, use.names = FALSE),
    Std_Error = unlist(std_dev, use.names = FALSE)
  )
}

get_raw_latex_tab<- function(sde,show=TRUE) {

  estimates=as.list(sde$tmb_rep(),what="Est")
  std=as.list(sde$tmb_rep(),what="Std")

  # Create data frames for each set
  log_sigma_obs <- combine_to_df(estimates$log_sigma_obs, 
                                 std$log_sigma_obs, "log_sigma_obs")
  coeff_fe <- combine_to_df(estimates$coeff_fe,
                            std$coeff_fe, 
                            rownames(estimates$coeff_fe))

  coeff_re <- combine_to_df(estimates$coeff_re,
                            std$coeff_re, 
                            rownames(estimates$coeff_re))

  log_lambda <- combine_to_df(estimates$log_lambda,
                            std$log_lambda, 
                            rownames(estimates$log_lambda))

  latex_table <- xtable(table_data, 
                        caption = "Parameter Estimates and Standard Errors")

  if (show) {
    print(latex_table, type = "latex", include.rownames = FALSE)
  }

  return (latex_table)

@@

```{r estimates baseline_ctcrw0}
raw_tab_baseline_ctcrw0=get_raw_latex_tab(baseline_ctcrw0)
```

We can get confidence intervals for the intercept of each parameter and the random effects standard deviations.

```{r define function to get final results}

get_final_baseline_tab<-function(sde,include_sigma_obs=FALSE,show=TRUE) {

  post_coeff=sde$post_coeff(n_post=10000)
  post_par=list(
  "tau"=exp(post_coeff$coeff_fe[,"tau.(Intercept)"]),
  "nu"=exp(post_coeff$coeff_fe[,"nu.(Intercept)"]),
  "sigma_tau"=1/sqrt(exp(post_coeff$log_lambda[,"tau.s(ID)"])),
  "sigma_nu"=1/sqrt(exp(post_coeff$log_lambda[,"nu.s(ID)"])))

  if (include_sigma_obs) {
    post_par$sigma_obs=1000*exp(post_coeff$log_sigma_obs)
  }

  mean_par=lapply(post_par,mean)
  sd_par=lapply(post_par,sd)
  quant_par <- lapply(post_par,
                      quantile, probs = c(0.025, 0.975))

  parameter_names <- names(mean_par)
  estimates <- mean_par
  conf_intervals <- sapply(parameter_names, function(param) {
  sprintf("$[%.2f; %.2f]$", quant_par[[param]][1], 
          quant_par[[param]][2])
    })

  parameter_names=c("$\\tau$", "$\\nu$","$\\sigma_{\\tau}$","$\\sigma_{\\nu}$")
  if (include_sigma_obs) {
    parameter_names=c(parameter_names,"$\\sigma_{\\obs}$")
  }
  table_data <- data.frame(
    Parameter = parameter_names,
  Estimate = sprintf("$%.2f$", mean_par),
  CI = conf_intervals,
  stringsAsFactors = FALSE)

  xtab <- xtable(
  table_data,
  caption = paste(sde$type(),"baseline estimations with random effects"),
  label = paste0("table:",sde$type(),"_baseline_estimations_with_random_effects"))

  if (show) {
  print(
  xtab,type="latex",
  include.rownames = FALSE,
  sanitize.text.function = identity,  # Prevent escaping of LaTeX math symbols
  hline.after = c(-1, 0, nrow(table_data)),  
  table.placement="H")
  }

  return(xtab)

@@

```{r final estimates baseline_ctcrw0, message=FALSE}

final_tab_baseline_ctcrw0=get_final_baseline_tab(baseline_ctcrw0,include_sigma_obs=FALSE)

@@

@@

We now decide to introduce a rotational parameter $\omega$ in the SDE to allow the velocity to rotate depending on the narwhals position with respect to the shoreline.
We start by introducing the covariate \texttt{theta} in the new smooth parameter $\omega$ to understand how the animals may rotate depending on the direction of their movement with respect to the shoreline.

@@

#model formula
formulas <- list(mu1=~1,mu2=~1,
                 tau =~s(theta_translated,k=10,bs="cc")+s(ID,bs="re"),
                 nu=~s(ID,bs="re"),
                 omega=~s(theta_translated,k=5,bs="cs"))

# Fit baseline with measurement error estimated from the data
baseline_racvm0<- SDE$new(formulas = formulas,data = dataBE12,type = "RACVM_SSM",response = c("x","y"),
                    par0 = par0,other_data=list("H"=H),fixpar=c("mu1","mu2"))

@@

```{r estimates baseline_racvm0}
raw_tab_baseline_racvm0=get_raw_latex_tab(baseline_racvm0)
```

```{r final estimates baseline_racvm0, message=FALSE}

final_tab_baseline_racvm0=get_final_baseline_tab(baseline_racvm0,include_sigma_obs=FALSE)

@@

@@

We add the effect of the distance to the shore through the covariate \texttt{Ishore_translated}.
This model has a lot of parameters and takes around 15 minutes to fit. 
Indeed, we will need to fix the penalties in the tensor splines in advance to ensure convergence of the optimizer. The standard deviation of the random effects and the smoothing penalties for the splines in $\tau$ will still be estimated from the data and not fixed. To illustrate this we first fit a model where the smoothing penalties are all estimated from the data.

@@

#parameters formulas
formulas <- list(mu1 = ~1 ,mu2 =~1,tau =~s(theta_translated,k=10,bs="cc")+s(ID,bs="re"),nu=~s(ID,bs="re"),
                   omega=~te(theta_translated,Ishore_translated,k=5,bs="cs"))

@@

```{r std random effects baseline_racvm1_unstable}

@@

The smoothing penalties estimates are extremely low (the smoothing penalties are the inverse of the square root of \texttt{omega.te(theta\_translated,Ishore\_translated)}). This suggests that the optimization did not go well and numerical errors occurred. For this reason, we decide to fix the smoothing penalties to "reasonable" values in the tensor splines.

First, we get good values for the smoothing penalty by approximating an expected "reasonable" surface with tensor splines, using the parametric function defined in the paper.

@@

fomega=function(cov_data,a=1,b=2,D0=0.3,D1=1,sigma_D=0.3,sigma_theta=pi/6){
  Dshore=cov_data$DistanceShore
  theta=cov_data$theta
  if (is.null(Dshore)){
    Dshore=1/cov_data$Ishore
  }

@@

# Splines smooth for omega -------------------- 
n <- 100000
SP_DF=c(5,5)

#sample theta and DistanceShore points in the domain
theta <- runif(n,-pi,pi);
DistanceShore <- runif(n,0.05,3)
samples=data.frame(theta=theta,DistanceShore=DistanceShore)

#define grid of values
theta_v <- seq(-pi,pi,length=30)
Dshore_v<- seq(0.05,3,length=30)
pr <- data.frame(theta=rep(theta_v,30),DistanceShore=rep(Dshore_v,rep(30,30)))

@@

Ishore=1/DistanceShore

#fit with bivariate splines te
m1 <- gam(y~te(theta,Ishore,k=SP_DF,bs=c("cs")))

@@

#initial parameters
par0 <- c(0,0,1,4,0)
init_lambda=c(exp(3),exp(3),1,m1$sp)

@@

#parameters formulas
formulas <- list(mu1 = ~1 ,mu2 =~1,tau =~s(theta_translated,k=10,bs="cc")+s(ID,bs="re"),nu=~s(ID,bs="re"),
                   omega=~te(theta_translated,Ishore_translated,k=c(5,5),bs="cs"))

# Fit baseline
baseline_racvm1<- SDE$new(formulas = formulas,data = dataBE12,type = "RACVM_SSM",
                    response = c("x","y"),par0 = par0,other_data=list("H"=H),
                    fixpar=c("mu1","mu2"))

@@

```{r estimates baseline_racvm1}
raw_tab_baseline_racvm1=get_raw_latex_tab(baseline_racvm1)
```

```{r final estimates baseline_racvm1, message=FALSE}

final_tab_baseline_racvm1=get_final_baseline_tab(baseline_racvm1,include_sigma_obs=FALSE)

@@

```{r display plots baseline_racvm1}

Ishore_values=c(1/0.25,1/2,1/10)
plots=lapply(1:length(Ishore_values),function(i){
  baseline_racvm1$plot_par(var="theta_translated",covs=data.frame("ID"="Asgeir","Ishore_translated"=Ishore_values[i]),
                         par_names=c("omega"),n_post=100)})
plots

@@

xmin=list("Ishore_translated"=1/3)
xmax=list("Ishore_translated"=1/0.3)
link=list("Ishore_translated"=\(x) 1/x)
xlabel=list("Ishore_translated"="DistanceShore")

plots_bas_racvm1=baseline_racvm1$get_all_plots(baseline=NULL,
      xmin=xmin,xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise")

@@

```{r save plots baseline_racvm1}
# Create directory
if (!dir.exists("baseline_racvm1")) {
    dir.create("baseline_racvm1", recursive = TRUE)
}

# Use lapply to save each plot
lapply(seq_along(plots_bas_racvm1), function(i) {

    plot<-plots_bas_racvm1[[i]]
    plot_name <- names(plots_bas_racvm1)[i]
    file_path <- file.path("baseline_racvm1", plot_name)

    # Check if the plot is a ggplot or plotly object
    if (inherits(plot, "ggplot")) {
      ggsave(file_path, plot, device = "pdf")
    } else if (inherits(plot, "plotly")) {
      saveWidget(plot, file_path)
  }
    })
```
Now, we would like to demonstrate how the Distance to shore influences the smooth parameter $\omega$.
We plot the estimated smooth parameter $\omega$ as a function of the angle for different values of the Distance to shore.

```{r smooth omega baseline_racvm1}

get_shore_influence_plot<-function(sde) {

  xmin=list("Ishore_translated"=1/2)
  xmax=list("Ishore_translated"=1/0.2)
  link=list("Ishore_translated"=\(x) 1/x)
  xlabel=list("Ishore_translated"="DistanceShore")

  plots=sde$get_all_plots(baseline=NULL,
      xmin=xmin,xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise")

  plot1=plots$fe_omega_theta_translated_q1_Ishore_translated
  plot2=plots$fe_omega_theta_translated_q2_Ishore_translated
  plot4=plots$fe_omega_theta_translated_q4_Ishore_translated

@@

  combined_plot <- ggplot() +
    geom_line(data=combined_omega, aes(x=x, y=y, color=Category, linetype=Category)) +
    geom_ribbon(data=combined_ci_omega, aes(x=x, ymin=ymin, ymax=ymax, fill=Category), alpha=0.1) +
    labs(color="Category", linetype="Category", x=expression(Theta), y=expression(omega)) +
    geom_vline(xintercept=c(-pi/2, pi/2), linetype="dashed", color="grey") +
    annotate("text", x = -pi/2-0.3, y = -5, label = expression(-pi/2), color = "red", size = 8) + 
    annotate("text", x = pi/2+0.2, y = -5, label = expression(pi/2), color = "red", size = 8) + 
    theme_minimal()+
   theme(
    axis.title.x = element_text(size=22),
    axis.title.y = element_text(size=22, angle=0, vjust=0.5, margin = margin(r = 10)),
    axis.text.x = element_text(size=18),
    axis.text.y = element_text(size=18),
    legend.title = element_text(size=18),
    legend.text = element_text(size=18)
  )

  return(combined_plot)
}

shore_influence_plot_baseline_racvm1=get_shore_influence_plot(baseline_racvm1)

ggsave("omega_DistanceShore_levels.pdf", plot=shore_influence_plot_baseline_racvm1, path="baseline_racvm1", width=10, height=6, units="in")

@@
